Verification of Stencil Computations through Spatial Specifications

### 2016-07-06 paper.tex
actually its line 5 that swaps a and b...
fixed a typo
typos

\emph{Stencils} are a ubiquitous programming pattern, common in
scientific and numerical computing applications. Informally, a stencil
computation computes an array value, where the value at each index $i$ of
this array is calculated from a \emph{neighbourhood} of values around $i$ in
some input array(s), \eg{}, the Game of Life, convolutions in image
processing, approximations to differential equations. For example, the
following iteration computes the one-dimensional discrete Laplace
transform (an approximation to a derivative) in Fortran:
%
\begin{minted}{fortran}
do iter = 0, itermax
   do i = 1, (n-1)
      b(i) = a(i-1) - 2*a(i) + a(i+1)
   end do
   a = b
end do
\end{minted}
%
Line 3 is the core of the stencil computation, calculating
the value at \texttt{b(i)} from a neighbourhood of elements about
\texttt{a(i)}. Line 4 swaps
\texttt{a} and \texttt{b} between iterations, where \texttt{b} becomes the
input for the next iteration. In Fortran, parentheses \texttt{( )} are used
for array subscripts rather than the more familiar bracket syntax \texttt{[ ]}.

@@

Region specifications can be assigned to region variables via
a region declaration, which can be used later to form a spatial
specification. For example, the specification from the previous
 example can be restated as:
%%
\begin{minted}{fortran}
!= region r1 = centered(depth=1, dim=1)
!= region r2 = centered(depth=1, dim=2)
!= region robertsCross = r1*r2
!= stencil robertsCross :: a
\end{minted}
This is especially useful for common stencils such as Robert's cross
as the region can be defined once and used by region name many times.
%%
\paragraph{Spatial modifiers}
%%
There are two modifiers one used at specification the other at region
level to provide better descriptions for stencil computations.

At the specification level, the \texttt{readOnce} modifier enforces that no
index appears more than once. For example, in all of the previous examples the
\texttt{readOnce} modifier could be added, \eg{}
%
\begin{minted}{fortran}
!= stencil readOnce, backward(depth=2, dim=1) :: a
b(i+1) = a(i) + a(i-1) + a(i-2)
\end{minted}
%
However, this would be an invalid specification if any of the
array subscripts was repeated. This modifier provides a way to
rule out any accidental repetition of array subscripts.
The notion corresponds to that of \emph{linearity} in type systems,
though we opt for the more informative and easily understood name
\texttt{readOnce}. This modifier is optional, so it need not
be present even if the stencil is linear.

@@

\paragraph{Overlapping regions with +}
The first nine
equations (with labels of the form $\ldots \texttt{+} \ldots$)
explain the notion of \emph{overlapping} regions with sum
\term{+}. Note that in the first six, $r$ and $s$ range
over booleans denoting the absence (true) or presence (false)
of \irrefl{} which are combined via disjunction. For example,
the rule $(\textsc{B} \, \term{+} \,
\textsc{F})$ explains that the addition of \term{backward}
and \term{forward} regions in the same dimension and the same
depth is equivalent to a \term{centered} region of that depth.
If the \term{forward} part is irrefleixve (\ie{}, $r =
\textit{false}$) but \term{backward} is not (\ie{}, $s =
\textit{true}$) then their addition is equal to a \term{centered}
region with no \irrefl{} modifier by the disjunction ($r \vee s =
\textit{true}$), \ie{}:
%%
\begin{align*}
& \stenCenS{n}{d}  \\[-0.2em]
\equiv \;\; & \stenFwdSR{n}{d}{\irreflS} \; \texttt{+} \; \stenBwdS{n}{d}
\end{align*}
%%
That is, the \term{fwd} region is irreflexive describing code which is missing
a zero-offset neighbourhood index (\eg{}, missing $\texttt{a(i)}$), but the addition of
the \term{bwd} has no irreflexive attribute and thus provides the
missing zero offset index.

@@

Figure~\ref{fig:inequations} defines a notion of sub-specification
via the relation $<:^r$ on \textit{region} 
syntax and then $<:$ on \textit{spec} syntax inductively. 
Both relations are reflexive and transitive (we omit 
these equations for $<:$ on \textit{spec}), and congruent
with respect to \term{+}, \term{*}, \term{atLeast}, \term{atMost},
and \term{readOnce}. The (\trule{\textsc{eq}}) axiom connects
equations with inequations. The (\trule{\textsc{over}}) explains
the notion of spatial over approximation via overlapping regions.
The (\trule{\textsc{rule}}) allows the modifier that
enforces that every index is read only once to be dropped
as an approxiamtion. The (\trule{\textsc{bound}}) gives
a connection between the \term{atLeast} and \term{atMost}
approximation modifiers.

@@

We define a semantic model for our stencil specification language.
This model has a number of purposes: (1) it is used in the inference,
checking, and program synthesis algorithms (introduced in
Section~\ref{sec:analysis}); (2) gives a basis for the equational
theory and approximations of the previous section; and (3) can be used
to guide correct implementations. The model is a denotational
semantics, in terms of sets of vectors which we call \emph{index
  schemes}. We introduces these first then define the denotational semantics.

@@

Indexing schemes give an abstract
representation of index expressions in the source language.
Our model represents the meaning of stencil specifications
as indexing schemes, which explain the range of possible
indexing behaviours allowed under the specification.
The relationship between indexing schemes and source language
terms is given precisely by the partial function, \textsf{schematic}, which
maps syntactic array index terms to indexing schemes. This
operation forms the basis of the analysis required for the inference
and the checking procedures in Section~\ref{sec:analysis}.

@@

\begin{example}
Three array indexing terms
$(i, j+1)$, $(1, j+1)$ and $(3, j)$
are mapped by \textsf{schematic} to the 
index schemes: $\vtwoh{0}{1}$, $\vtwoh{\infty}{1}$
and $\vtwoh{\infty}{0}$ respectively. The following
relationship of generalises then holds:
%\[
%\xymatrix@R=0.2em@C=2em{
%[\infty \;\; 0] \ar@{-}[dr]^{\preceq_n} &  \\
%& [0 \;\; 1]  \\
%[\infty \;\; 1]  \ar@{-}[ur]_{\preceq_n} & 
%}
%\]
%%
\begin{equation*}
\vtwoh{\infty}{0} \; \not\preceq_2 \; \vtwoh{0}{1} 
\qquad
\vtwoh{\infty}{1} \; \preceq_2 \; \vtwoh{0}{1} 
\end{equation*}
%%
\end{example}
\noindent
Our terminology of \emph{indexing schemes} and
\emph{generalisation} are by analogy
with the \emph{type schemes} and \emph{type generalisation} of
polymorphic types in ML~\cite{milner1978theory}. Our notion is
much more rigid however, as our schemes do not bind names
and thus generalisation of does not require substitution; $\infty$ acts
as a positional placeholder.

@@

The pairwise permutation $u \bowtie v$ takes two vectors and builds the matrix
where the rows are all possible $n$-vectors generated by
non-deterministic picking each $i^{th}$ entry from either the
$i^{th}$ entry in $u$ or the $i^{th}$ entry in $j$. For example:
%
\begin{equation*}
\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} =
\left[
\setlength{\arraycolsep}{0.42em}
\begin{array}{cccccccc}
0 & 0 & 0 & 0 & 3 & 3 & 3 & 3 \\ 
1 & 1 & 4 & 4 & 1 & 1 & 4 & 4 \\
2 & 5 & 2 & 5 & 2 & 5 & 2 & 5
\end{array}
\right]
% \\
%\setlength{\arraycolsep}{0.5em}
%\left[\begin{array}{ccc}
%0 & 1 & 2 \\
%0 & 1 & 5 \\
%0 & 4 & 2 \\
%0 & 4 & 5 \\
%3 & 1 & 2 \\
%3 & 1 & 5 \\
%3 & 4 & 2 \\
%3 & 4 & 5
%\end{array}
%\right]
\end{equation*}
%
The $2^n$ unique choices for $\bowtie$ on $n$-vectors
corresponds to taking all bit-strings of length $n$ and
selecting from $u$ for $1$ and $v$ for $0$. The pairwise permutation
matrix  s thus defined in terms of the logical
matrix $b$ where $b_{i,j}$ is the $j$-th bit of the integer
$i - 1$ as:
%
\begin{align*}
(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
\end{align*}
%
\end{definition}

>>>>

\emph{Stencils} are a ubiquitous programming pattern, common in
scientific and numerical computing applications. Informally, a stencil
computation computes an array value, where the value at each index $i$ of
this array is calculated from a \emph{neighbourhood} of values around $i$ in
some input array(s), \eg{}, the Game of Life, convolutions in image
processing, approximations to differential equations. For example, the
following iteration computes the one-dimensional discrete Laplace
transform (an approximation to a derivative) in Fortran:
%
\begin{minted}{fortran}
do iter = 0, itermax
   do i = 1, (n-1)
      b(i) = a(i-1) - 2*a(i) + a(i+1)
   end do
   a = b
end do
\end{minted}
%
Line 3 is the core of the stencil computation, calculating
the value at \texttt{b(i)} from a neighbourhood of elements about
\texttt{a(i)}. Line 5 swaps
\texttt{a} and \texttt{b} between iterations, where \texttt{b} becomes the
input for the next iteration. In Fortran, parentheses \texttt{( )} are used
for array subscripts rather than the more familiar bracket syntax \texttt{[ ]}.

@@

Region specifications can be assigned to region variables via
a region declaration, which can be used later to form a spatial
specification. For example, the specification from the previous
 example can be restated as:
%%
\begin{minted}{fortran}
!= region r1 = centered(depth=1, dim=1)
!= region r2 = centered(depth=1, dim=2)
!= region robertsCross = r1*r2
!= stencil robertsCross :: a
\end{minted}
This is especially useful for common stencils such as Robert's cross
as the region can be defined once and used by region name many times.
%%
\paragraph{Spatial modifiers}
%%
There are two modifiers: one used at specification and the other at region
level to provide better descriptions for stencil computations.

At the specification level, the \texttt{readOnce} modifier enforces that no
index appears more than once. For example, in all of the previous examples the
\texttt{readOnce} modifier could be added, \eg{}
%
\begin{minted}{fortran}
!= stencil readOnce, backward(depth=2, dim=1) :: a
b(i+1) = a(i) + a(i-1) + a(i-2)
\end{minted}
%
However, this would be an invalid specification if any of the
array subscripts were repeated. This modifier provides a way to
rule out any accidental repetition of array subscripts.
The notion corresponds to that of \emph{linearity} in type systems,
though we opt for the more informative and easily understood name
\texttt{readOnce}. This modifier is optional, so it need not
be present even if the stencil is linear.

@@

\paragraph{Overlapping regions with +}
The first nine
equations (with labels of the form $\ldots \texttt{+} \ldots$)
explain the notion of \emph{overlapping} regions with sum
\term{+}. Note that in the first six, $r$ and $s$ range
over booleans denoting the absence (true) or presence (false)
of \irrefl{}. These are combined via disjunction. For example,
the rule $(\textsc{B} \, \term{+} \,
\textsc{F})$ explains that the addition of \term{backward}
and \term{forward} regions in the same dimension and the same
depth is equivalent to a \term{centered} region of that depth.
If the \term{forward} part is irrefleixve (\ie{}, $r =
\textit{false}$) but \term{backward} is not (\ie{}, $s =
\textit{true}$) then their addition is equal to a \term{centered}
region with no \irrefl{} modifier by the disjunction ($r \vee s =
\textit{true}$), \ie{}:
%%
\begin{align*}
& \stenCenS{n}{d}  \\[-0.2em]
\equiv \;\; & \stenFwdSR{n}{d}{\irreflS} \; \texttt{+} \; \stenBwdS{n}{d}
\end{align*}
%%
That is, the \term{fwd} region is irreflexive describing code which is missing
a zero-offset neighbourhood index (\eg{}, missing $\texttt{a(i)}$), but the addition of
the \term{bwd} has no irreflexive attribute and thus provides the
missing zero offset index.

@@

Figure~\ref{fig:inequations} defines a notion of sub-specifications
via the relation $<:^r$ on \textit{region} 
syntax and then $<:$ on \textit{spec} syntax inductively. 
Both relations are reflexive and transitive (we omit 
these equations for $<:$ on \textit{spec}), and congruent
with respect to \term{+}, \term{*}, \term{atLeast}, \term{atMost},
and \term{readOnce}. The (\trule{\textsc{eq}}) axiom connects
equations with inequations. The (\trule{\textsc{over}}) explains
the notion of spatial over approximation via overlapping regions.
The (\trule{\textsc{rule}}) allows the modifier that
enforces that every index is read only once to be dropped
as an approxiamtion. The (\trule{\textsc{bound}}) gives
a connection between the \term{atLeast} and \term{atMost}
approximation modifiers.

\anote{Can we say something here about why these equations are useful i.e. we use them to do simplification of stencils?}

@@

We define a semantic model for our stencil specification language.
This model has a number of purposes: (1) it is used in the inference,
checking, and program synthesis algorithms (introduced in
Section~\ref{sec:analysis}); (2) gives a basis for the equational
theory and approximations of the previous section; and (3) can be used
to guide correct implementations. The model is a denotational
semantics, in terms of sets of vectors which we call \emph{index
  schemes}. We introduce these first and then define the denotational semantics.

@@

Indexing schemes give an abstract
representation of index expressions in the source language.
Our model represents the meaning of stencil specifications
as indexing schemes, which explain the range of possible
indexing behaviours allowed under the specification.
The relationship between indexing schemes and source language
terms is given precisely by the partial function, \textsf{schematic}, which
maps syntactic array-index terms to indexing schemes. This
operation forms the basis of the analysis required for the inference
and the checking procedures in Section~\ref{sec:analysis}.

@@

\begin{example}
Three array indexing terms
$(i, j+1)$, $(1, j+1)$ and $(3, j)$
are mapped by \textsf{schematic} to the 
index schemes: $\vtwoh{0}{1}$, $\vtwoh{\infty}{1}$
and $\vtwoh{\infty}{0}$ respectively. The following
relationship of generalises then holds:
%\[
%\xymatrix@R=0.2em@C=2em{
%[\infty \;\; 0] \ar@{-}[dr]^{\preceq_n} &  \\
%& [0 \;\; 1]  \\
%[\infty \;\; 1]  \ar@{-}[ur]_{\preceq_n} & 
%}
%\]
%%
\begin{equation*}
\vtwoh{\infty}{0} \; \not\preceq_2 \; \vtwoh{0}{1} 
\qquad
\vtwoh{\infty}{1} \; \preceq_2 \; \vtwoh{0}{1} 
\end{equation*}
%%
\end{example}
\noindent
Our terminology of \emph{indexing schemes} and
\emph{generalisation} is by analogy
with the \emph{type schemes} and \emph{type generalisation} of
polymorphic types in ML~\cite{milner1978theory}. Our notion is
much more rigid however, as our schemes do not bind names
and thus generalisation does not require substitution; $\infty$ acts
as a positional placeholder.

@@

The pairwise permutation $u \bowtie v$ takes two vectors and builds the matrix
where the rows are all possible $n$-vectors generated by
non-deterministic picking each $i^{th}$ entry from either the
$i^{th}$ entry in $u$ or the $i^{th}$ entry in $j$. For example:
%
\begin{equation*}
\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} =
\left[
\setlength{\arraycolsep}{0.42em}
\begin{array}{cccccccc}
0 & 0 & 0 & 0 & 3 & 3 & 3 & 3 \\ 
1 & 1 & 4 & 4 & 1 & 1 & 4 & 4 \\
2 & 5 & 2 & 5 & 2 & 5 & 2 & 5
\end{array}
\right]
% \\
%\setlength{\arraycolsep}{0.5em}
%\left[\begin{array}{ccc}
%0 & 1 & 2 \\
%0 & 1 & 5 \\
%0 & 4 & 2 \\
%0 & 4 & 5 \\
%3 & 1 & 2 \\
%3 & 1 & 5 \\
%3 & 4 & 2 \\
%3 & 4 & 5
%\end{array}
%\right]
\end{equation*}
%
The $2^n$ unique choices for $\bowtie$ on $n$-vectors
corresponds to taking all bit-strings of length $n$ and
selecting from $u$ for $1$ and $v$ for $0$. The pairwise permutation
matrix is thus defined in terms of the logical
matrix $b$ where $b_{i,j}$ is the $j$-th bit of the integer
$i - 1$ as:
%
\begin{align*}
(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
\end{align*}
%
\end{definition}
